---
title: 'Review of Data Science 1'
format:
  html: default
editor_options: 
  chunk_output_type: console
---

You can download this .qmd file from [here](https://github.com/proback/264_fall_2026/blob/main/01_review164.qmd).  Just hit the Download Raw File button.


### Determinants of COVID vaccination rates

```{r}
install.packages("tidyverse")
install.packages("credentials")
```

```{r}
#| include: false

library(tidyverse)
vaccine_data <- read_csv("https://proback.github.io/264_fall_2026/Data/vaccinations_2021.csv")
```

First, a little detour to describe several alternatives for reading in data:

If you navigate to [my Github account](https://github.com/proback), and find the `264_fall_2026` repo, there is a Data folder inside.  You can then click on `vacc_Mar21.csv` to see the data we want to download.   [This link](https://github.com/proback/264_fall_2026/blob/main/Data/vaccinations_2021.csv) should also get you there, but it's good to be able to navigate there yourself.


```{r}
#| eval: false

# Approach 1
vaccine_data <- read_csv("Data/vaccinations_2021.csv")  #<1>

# Approach 2
vaccine_data <- read_csv("~/264_fall_2026/Data/vaccinations_2021.csv")  #<2>

# Approach 3
vaccine_data <- read_csv("https://proback.github.io/264_fall_2026/Data/vaccinations_2021.csv")  #<3>

# Approach 4
vaccine_data <- read_csv("https://raw.githubusercontent.com/proback/264_fall_2026/refs/heads/main/Data/vaccinations_2021.csv")  #<4>

```

1. Approach 1: create a Data folder in the same location where this .qmd file resides, and then store vaccinations_2021.csv in that Data folder
2. Approach 2: give R the complete path to the location of vaccinations_2021.csv, starting with Home (~)
3. Approach 3: link to our course webpage, and then know we have a Data folder containing all our csvs
4. Approach 4: navigate to the data in GitHub, hit the Raw button, and copy that link


A recent Stat 272 project examined determinants of covid vaccination rates at the county level.  Our data set contains 3053 rows (1 for each county in the US) and 14 columns; here is a quick description of the variables we'll be using:

- `state` = state the county is located in
- `county` = name of the county
- `region` = region the state is located in
- `metro_status` = Is the county considered "Metro" or "Non-metro"?
- `rural_urban_code` = from 1 (most urban) to 9 (most rural)
- `perc_complete_vac` = percent of county completely vaccinated as of 11/9/21
- `tot_pop` = total population in the county
- `votes_Trump` = number of votes for Trump in the county in 2020
- `votes_Biden` = number of votes for Biden in the county in 2020
- `perc_Biden` = percent of votes for Biden in the county in 2020
- `ed_somecol_perc` = percent with some education beyond high school (but not a Bachelor's degree)
- `ed_bachormore_perc` = percent with a Bachelor's degree or more
- `unemployment_rate_2020` = county unemployment rate in 2020
- `median_HHincome_2019` = county's median household income in 2019


1. Consider only Minnesota and its surrounding states (Iowa, Wisconsin, North Dakota, and South Dakota).  We want to examine the relationship between the percentage who voted for Biden and the percentage of complete vaccinations by state.  Generate two plots to examine this relationship:

a) A scatterplot with points and smoothers colored by state.  Make sure the legend is ordered in a meaningful way, and include good labels on your axes and your legend.  Also leave off the error bars from your smoothers.

```{r}
vaccine_data |>
  filter(state %in% c("Minnesota", "Wisconsin", "North Dakota", "South Dakota", "Iowa")) |>
  ggplot(aes(x = perc_Biden, y = perc_complete_vac, color = fct_reorder2(state, perc_Biden, perc_complete_vac))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Percentage of Biden Voters vs. Percentage Vaccinated by State",
       x = "Percentage of Biden Voters",
       y = "Percentage Vaccinated",
       color = "State")
```

b) One plot per state containing a scatterplot and a smoother.
```{r}
vaccine_data |>
  filter(state %in% c("Minnesota", "Wisconsin", "North Dakota", "South Dakota", "Iowa")) |>
  ggplot(aes(x = perc_Biden, y = perc_complete_vac)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~state)
```

Describe which plot you prefer and why.  What can you learn from your preferred plot?
**The second plot because it is easier to see. Only benefit to the first one is that it is easier to compare (but it is messier)**

4. Produce 3 different plots for illustrating the relationship between the rural_urban_code and percent vaccinated.  Hint: you can sometimes turn numeric variables into categorical variables for plotting purposes (e.g. `as.factor()`, `ifelse()`).

```{r}
vaccine_data_rsu <- vaccine_data |>
  mutate(rural = ifelse(rural_urban_code < 5, "More Urban", "More Rural"))

vaccine_data_rsu |>
  ggplot(aes(x = rural_urban_code, y = perc_complete_vac, color = rural_urban_code)) +
  geom_jitter() +
  geom_smooth(method = "lm", se = FALSE)

vaccine_data_rsu |>
  ggplot(aes(x = as_factor(rural_urban_code), fill = perc_complete_vac > 50)) +
  geom_bar(position = "fill")

vaccine_data_rsu |>
  ggplot(aes(x = rural_urban_code, y = perc_complete_vac)) +
  geom_point()
```

State your favorite plot, why you like it better than the other two, and what you can learn from your favorite plot.  Create an alt text description of your favorite plot, using the Four Ingredient Model.  See [this link](https://drive.google.com/file/d/1BuEbL6vETBcFULmtlrwwZBUYhdEg0Zqb/view?usp=sharing) for reminders and references about alt text.


5. BEFORE running the code below, sketch the plot that will be produced by R.  AFTER running the code, describe what conclusion(s) can we draw from this plot?  

```{r}
#| eval: FALSE

vaccine_data |>
  filter(!is.na(perc_Biden)) |> # removes n/a values
  mutate(big_states = fct_lump(state, n = 10)) |> # leaves 10 states with the most counties, lumps everything else into "other"
  group_by(big_states) |> # groups by big_states
  summarize(IQR_Biden = IQR(perc_Biden)) |> # Finds interquartile ranges for each of the 11 groups
  mutate(big_states = fct_reorder(big_states, IQR_Biden)) |> # orders big_states by iqr_biden
  ggplot() + 
    geom_point(aes(x = IQR_Biden, y = big_states)) # will make a scatterplot with big_states on the y-axis and IQR_Biden on the x-axis, with groups on the y-axis aligned by ascending values of IQR_Biden
# only one data point per group
```

This plot shows the political variation based on the interquartile range of percentage of Biden votes (so it ignores extreme values). This is a plot that details the 10 states with the most counties and all other states lumped together as one data point. We see that, of the groups detailed in this plot, Tennessee has the lowest variation while Virginia has the highest, closely followed by the "Other" group.

6. In this question we will focus only on the 12 states in the Midwest (i.e. where region == "Midwest").

a) Create a tibble with the following information for each state.  Order states from least to greatest state population.

- number of different `rural_urban_code`s represented among the state's counties (there are 9 possible)
- total state population
- proportion of Metro counties
- median unemployment rate 

```{r}
midwest_data <- vaccine_data |>
  filter(region == "Midwest")

midwest <- midwest_data |>
  group_by(state) |>
  summarize(
    num_ruc = n_distinct(rural_urban_code),
    pop = sum(tot_pop),
    prop_metro = mean(metro_status == "Metro"),
    med_ur = median(unemployment_rate_2020)
  ) |>
  arrange(pop)
  

midwest
```

b) Use your tibble in (a) to produce a plot of the relationship between proportion of Metro counties and median unemployment rate.  Points should be colored by the number of different `rural_urban_code`s in a state, but a single linear trend should be fit to all points.  What can you conclude from the plot?

```{r}
midwest |>
  ggplot(aes(x = prop_metro, y = med_ur, color = num_ruc)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

We can conclude that there is a somewhat strong positive linear relationship between the proportion of Metro counties and median unemployment rate. The number of distinct urban rural codes does not seem to have much influence. There are also two obvious outliers, being around the 0.3 Metro proportion range. One exists far above the line and the other below.


In this next section, we consider a few variables that could have been included in our data set, but were NOT.  Thus, you won't be able to write and test code, but you nevertheless should be able to use your knowledge of the tidyverse to answer these questions.

Here are the hypothetical variables:

- HR_party = party of that county's US Representative (Republican, Democrat, Independent, Green, or Libertarian)
- people_per_MD = number of residents per doctor (higher values = fewer doctors)
- perc_over_65 = percent of residents over 65 years old
- perc_white = percent of residents who identify as white


8. Hypothetical R chunk #1:

```{r}
#| eval: false

# Hypothetical R chunk 1
temp <- vaccine_data |>
  mutate(new_perc_vac = ifelse(perc_complete_vac > 95, NA, perc_complete_vac),
         MD_group = cut_number(people_per_MD, 3)) |>
  group_by(MD_group) |>
  summarise(n = n(),
            mean_perc_vac = mean(new_perc_vac, na.rm = TRUE),
            mean_white = mean(perc_white, na.rm = TRUE))
```  

a) Describe the tibble `temp` created above.  What would be the dimensions?  What do rows and columns represent?

b) What would happen if we replaced `new_perc_vac = ifelse(perc_complete_vac > 95, NA, perc_complete_vac)` with `new_perc_vac = ifelse(perc_complete_vac > 95, perc_complete_vac, NA)`?

c) What would happen if we replaced `mean_white = mean(perc_white, na.rm = TRUE)` with `mean_white = mean(perc_white)`?

d) What would happen if we removed `group_by(MD_group)`?


9. Hypothetical R chunk #2:

```{r}
#| eval: false

# Hypothetical R chunk 2
ggplot(data = vaccine_data) +
  geom_point(mapping = aes(x = perc_over_65, y = perc_complete_vac, 
                           color = HR_party)) +
  geom_smooth()

temp <- vaccine_data |>
  group_by(HR_party) |>
  summarise(var1 = n()) |>
  arrange(desc(var1)) |>
  slice_head(n = 3)

vaccine_data |>
  ggplot(mapping = aes(x = fct_reorder(HR_party, perc_over_65, .fun = median), 
                       y = perc_over_65)) +
    geom_boxplot()
```

a) Why would the first plot produce an error?

**Because mapping is only defined in the geom_point function and not in the geom_smooth function.**

b) Describe the tibble `temp` created above.  What would be the dimensions?  What do rows and columns represent?

**Temp is a 3 x 2 tibble with the first column listing the top 3 HR_parties with the most data points in descending order, and the second column would be var1, and would have the number of data points for each HR_party.**

c) What would happen if we replaced `fct_reorder(HR_party, perc_over_65, .fun = median)` with `HR_party`?

**The HR_parties would not be ordered by the median values of perc_over_65. The actual data would remain the same, but is a bit more disorganized than the current version.**

10. Hypothetical R chunk #3:

```{r}
#| eval: false

# Hypothetical R chunk 3
vaccine_data |>
  filter(!is.na(people_per_MD)) |>
  mutate(state_lump = fct_lump(state, n = 4)) |>
  group_by(state_lump, rural_urban_code) |>
  summarise(mean_people_per_MD = mean(people_per_MD)) |>
  ggplot(mapping = aes(x = rural_urban_code, y = mean_people_per_MD, 
      colour = fct_reorder2(state_lump, rural_urban_code, mean_people_per_MD))) +
    geom_line()
```

a) Describe the tibble piped into the ggplot above.  What would be the dimensions?  What do rows and columns represent?

b) Carefully describe the plot created above.

c) What would happen if we removed `filter(!is.na(people_per_MD))`?

d) What would happen if we replaced `fct_reorder2(state_lump, rural_urban_code, mean_people_per_MD)` with `state_lump`?